Copulas and their potential for ecology

### 2019-12-20 Paper.Rmd
eqn edited upto sec8 in maintext

Well-developed approaches do exist, however, and are 
applied widely in other fields, 
that provide a description of the association 
between two or more variables that is mathematically provably 
complete [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]. 
The approaches are based on 
the concept of the *copula*. Copula approaches separate information in a bivariate 
random variable into two parts: the information in the 
marginal distributions (which says nothing about the association between the 
variables), and the remaining information (which is solely about
the association). We here introduce some simple concepts, based on ranks,
that relate to copulas and give a conceptual 
flavor of copula ideas to be introduced 
formally in the "Background" section below (section \ref{Background}). Given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), information about the structure of the association
between $x$ and $y$ can be separated from information contained 
in the marginal distributions by 
considering the plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The rank
of the smallest element in a set is $1$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$; they relate to 
the empirical cumulative distribution function of the $x_i$ and $y_i$,
respectively. We henceforth abbreviate "cumulative distribution function" as "cdf". 
See, for instance,
Fig. \ref{fig_cop_pedag1}C-D, which show the normalized ranks of 
A-B. Ranking makes the marginal distributions of 
the component datasets uniform, isolating the 
dependence structure. Dependence structure and marginals can then be studied 
separately. Normalized rank plots such as Fig. \ref{fig_cop_pedag1}C-D
relate to copulas in a way to be described 
(section \ref{Background}). Note that Pearson correlation, even though it is the most commonly
used measure of association, is modified by monotonic
tranformation of the component variables (Appendix 
\ref{SM-pedagog1_details}, Fig. \ref{SM-fig_transf_pearson_spearman}),
and therefore reflects not only dependence information, but also
information on the marginals [@Genest2007]. 
Spearman and Kendall correlations, being rank-based, 
are not influenced by monotonic transformations of the variables. 
They provide information solely on the dependence between those variables.

@@

A second reason why we may expect detailed understanding of dependence structures
to benefit ecological
research is that prior work demonstrated, using copulas, complex 
stucture in the spatial dependence of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. An environmental variable 
measured through time in two locations may show strong tail associations between
the locations if
intense meteorological events are also widespread, as seems 
frequently to be the case: extreme values are associated with intense events,
so happen in both places at the same time, whereas moderate values of the 
environmental variable may be associated with local phenomena, which differ
between the locations.
Spatial dependence in an environmental variable 
tends to beget spatial dependence 
in fluctuations of populations or other ecosystem variables 
(this is called \emph{spatial synchrony}) influenced by the environmental 
variable. This is called the \emph{Moran effect}. 
If asymmetric tail associations, or other complex dependence structure, is transmitted 
from environmental to ecological variables, then we would expect complex dependence 
structure and tail associations to be a common feature of the 
spatial synchrony of population, community, 
biogeochemical and other environmentally influenced ecological variables.
Synchrony attracts major interest in 
ecology [@Liebhold2004; @Sheppard2016; @walteretal2017].

@@

We approached our overall goal of exploring the potential of copula approaches 
for ecology by addressing the following specific questions. (Q1) Do datasets
in ecology have dependence structure distinct from that of a multivariate 
Gaussian/normal
distribution (here called non-normal copula structure)? Do positively associated 
ecological variables show tail 
associations distinct from those of a normal distribution, and in particular
do they show asymmetric tail associations? Normal copula structure is assumed
by standard approaches that use multivariate 
normal distributions or distributions obtained by transforming
the marginals of a normal distribution. So Q1 asks whether ecological
data contain dependency information distinct from a standard or default case. 
(Q2) If the answer to
Q1 is "yes," then what are some possible causes/mechanisms of non-normal
copula structure and asymmetric tail associations in ecology? And (Q3), what are the 
consequences of non-normal copula
structure and asymmetric tail associations for ecological understanding and
applications? After section \ref{Background}, 'Background on copulas', we address Q1 
(sections \ref{Data} through \ref{Results} of the paper) 
by analyzing several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
We address Q2 (sections \ref{Causes} through \ref{Results_Q2}), using 
simple models. We address Q3 (sections \ref{Consequences} through \ref{Results_Q3}) using 
both data and models. Section \ref{Discussion} is the Discussion.
Multiple analyses are brought to bear on each question, and these
are summarized diagramatically in Fig. \ref{MasterFigure}, which can 
serve as a *post hoc* guide to the paper. 
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000],
and have also been used effectively, but rarely so far, in ecology [@Valpine2014; 
@Anderson2018; @Popovic2018]. 
We argue that benefits of wider use of
copula descriptions of dependence in ecology will be substantial.

@@

Plots using normalized ranks were proposed in the Introduction as being conceptually
similar to copulas; we explain the connection. As stated above, the copula associated
with a random variable $(X,Y)$ is the cdf of the random variable
$(F_X(X),F_Y(Y))$. Given a sample $(x_i,y_i)$, $i=1,\ldots,n$, from $(X,Y)$, let $\hat{F}_X$ and
$\hat{F}_Y$ be the empirical cdfs associated with the $x_i$ and $y_i$, respectively.
These are step functions of $x$ and $y$, respectively, that start at $0$ for low values of 
$x$ and $y$ and jump by $1/n$
at each of the data points. Therefore $\hat{F}_X(x_j)$ is the rank of $x_j$ in
the set $\{x_i : i=1,\ldots,n\}$, dividided by $n$, 
and analogously $\hat{F}_Y(y_j)$ is the rank of $y_j$ in the set 
$\{y_i : i=1,\dots,n\}$, divided
by $n$. For large $n$, $\hat{F}_X(x_i)$ and $\hat{F}_Y(y_i)$ 
approximate the cdfs $F_X$ and $F_Y$, respectively. 
But $\hat{F}_X(x_i)$ equals the normalized rank of $x_i$ times
$n/(n+1)$, and likewise $\hat{F}_Y(y_i)$ equals the normalized rank of $y_i$ times
$n/(n+1)$. So the normalized rank in turn approximates the empirical cdf and therefore 
the cdf. Thus the normalized rank pairs $(u_i,v_i)$ for $i=1,\ldots,n$ can be 
regarded as approximate samples from the random variable $(F_X(X),F_Y(Y))$, 
which is the random variable associated with the copula of $(X,Y)$.
The scatterplot of $v_i$ versus $u_i$ can be used to infer aspects of copula 
structure (see section \ref{Methods} below).

@@

Model selection methods give the *relative* support of several models, but do not 
indicate whether any of the models are an objectively good fit. To test that, 
we tested the goodness of fit of our AIC-best copula family using a 
bootstrapping procedure of @Wang2000 and @Genest2006,
implemented as `BiCopGofTest` in `VineCopula`. The procedure performed 
one test based on a Cramer-von Mises statistic and another based on a 
Kolmogorov-Smirnov statistic. To keep computation times reasonable, 
a run using $100$ bootstraps was performed. If
the $p$-value from either test was $<0.2$, tests were re-run with 
$1000$ boostraps.

We fit 16 bivariate copula families, exhibiting a variety of 
lower- and upper-tail dependence characteristics, with bivariate datasets. The purpose
of using a large collection of families was to include a 
variety of alternative dependence structures to have a robust model
selection and multi-model inference procedure. For that purpose, it 
is not important for the reader to
understand the details of these copulas,
and, additonally, these copulas have been 
described in detail elsewhere [@Joe1997; @Brechmann2013]. So 
we only briefly identify each family, say a few words about its tail 
dependence, and provide pictorial descriptions (Figs 
\ref{figped_N}-\ref{figped_BB1}, \ref{SM-figped_SC} and \ref{SM-figped_G}-\ref{SM-figped_SBB8}). 
The pictorial decriptions can also be used to aid quick comparisons of copula
families when alternative families are being considered for future applications. 
We used several families that can exhibit positive lower-tail dependence 
(of strength depending on parameters) and 
zero upper-tail dependence: the Clayton, survival Gumbel, survival Joe
and survival BB6 families. We used several families that can exhibit positive
upper-tail dependence (strength depends on parameters) and zero lower-tail
dependence: the survival Clayton, Gumbel, Joe and BB6 families. We used
families that show zero upper- and lower-tail dependence: the normal
and Frank families. These families have pdfs symmetric about the line
$v=-u+1$. We also used several families that can show both upper- and
lower-tail dependence, in relative amounts depending on parameter
values: the BB1, survival BB1, BB7 and survival BB7 families. 
We also used the BB8 family, which shows zero lower-tail dependence,
and zero upper-tail dependence except for a boundary case for the parameters.
And we used the survival BB8 copula, which shows zero upper-tail dependence, 
and zero lower-tail dependence except for a boundary case.
"Survival" families are rotations of the copula with the similar name by
180 degrees. 
See @Joe1997 and @Brechmann2013 for details on all families. We used the implementations
of provided in the `VineCopula` package for the R programming
language. See Figs \ref{figped_N}-\ref{figped_BB1}, \ref{SM-figped_SC} and \ref{SM-figped_G}-\ref{SM-figped_SBB8} 
for visual depictions of the pdfs of these copulas and how
pdfs and tail dependence are influenced by the parameters. See 
Table \ref{tab_copula_info} for a summary.

@@

For large datasets (large $n$),
we used $l_b$ and $u_b$ close together without incurring undue
sampling variation in our statistics, and we considered 
multiple bands $(l_b,u_b)$ to understand how association
varies across different parts of the distributions. But for datasets with 
smaller $n$ we considered only $l_b=0$, $u_b=0.5$ and $l_b=0.5$, $u_b=1$,
abbreviating $\cor_l=\cor_{0,0.5}$ ($l$ is for "lower") 
and $\cor_u=\cor_{0.5,1}$ ($u$ is for "upper"). Likewise $\Ps_l=\Ps_{0,.5}$,
$\Ps_u=\Ps_{0.5,1}$, $\Dsq_l=\Dsq_{0,0.5}$, $\Dsq_u=\Dsq_{0.5,1}$.
To test for asymmetry of association in upper and lower portions of distributions,
we used differences $\cor_l-\cor_u$, $\Ps_l-\Ps_u$ and $\Dsq_u-\Dsq_l$
for smaller datasets (note the opposite order
in the last of these); and $\cor_{0,u_b}-\cor_{1-u_b,1}$, 
$\Ps_{0,u_b}-\Ps_{1-u_b,1}$ and
$\Dsq_{1-u_b,1}-\Dsq_{0,u_b}$ with $u_b$ close 
to $0$ for large datasets.
We tested our statistics in Appendix \ref{SM-Test_npa_stat} (see also 
Figs \ref{SM-fig_stat_testing35}-\ref{SM-fig_stat_testing1000}).

@@

The first mechanism relates to the ideas in the Introduction about 
Liebig's law of the minimum, and to nonlinear influences of environmental
variables on ecological variables.
If an environmental variable influences an ecological variable 
disproportionately in one of its tails, we explored whether the ecological
variable could then exhibit asymmetric tail associations across space.
Let $E_i(t)$ be an ecological variable 
measured at location $i$ ($i=1,2$) at time $t$.
Assume the dynamics 
$E_i(t+1)=bE_i(t)+g(\epsilon_i(t))+a \delta_i(t)$, where 
the $\delta_i(t)$ are standard-normally distributed 
and independent across time and locations, $a=0.2$,
$b=0.1, -0.1, 0.5, -0.5$ in different simulations, and 
the $(\epsilon_1(t),\epsilon_2(t))$ were drawn, independently through
time, from a bivariate normal distribution with $\var(\epsilon_i)=1$ and 
$\cov(\epsilon_1,\epsilon_2)=0.8$. Thus basic ecological dynamics follow the
very simple AR(1) formulation, influenced by a "regional" environmental factor which
is correlated across locations ($\epsilon$) and by a local factor ($\delta$). 
The function $g$ describes how $\epsilon_i(t)$ influences
$E_i(t)$. We used $g$ equal to $g_1$ or $g_2$ in different simulations:
$g_1(\epsilon)$ equals $\epsilon$ if $\epsilon<0$ and $0$ otherwise;
$g_2(\epsilon)$ equals $\epsilon$ if $\epsilon>0$ and $0$ otherwise. Thus
$g_1$ represents environmental effects that negatively impact populations,
but only below the threshold $\epsilon=0$; 
and $g_2$ represents effects that positively impact populations, but 
only above $\epsilon=0$.
The values of $b$ provide a modicum of exploration of whether ecological 
dynamics may also influence how complex dependencies between
the $E_i(t)$ arise: 
negative $b$ corresponds to overcompensating dynamics 
and positive $b$ to undercompensating dynamics; larger $|b|$
means slower return to equilibrium after a disturbance.

For each $b$ and $g_i$, we simulated the model for $25000$ time
steps and retained the $E_i(t)$ for the final $2500$ time 
steps. We applied our 
nonparametric statistic $\cor_{0,0.1}-\cor_{0.9,1}$ 
and our Spearman- or Kendall-preserving normal-copula 
surrogate comparison methods (see section \ref{Methods}) to these 
outputs to discover if the model could
produce asymmetric tail associations (and therefore non-normal copula structure)
between the $E_i(t)$. Because $(\epsilon_1,\epsilon_2)$
and $(\delta_1,\delta_2)$ have normal copula structure (they were drawn
from bivariate normal distributions), the Moran 
mechanism analyzed below does not operate here.

Our second mechanism is an extension of the well known Moran
effect, and was summarized conceptually in the Introduction. 
We consider a linear model, as well as 
two parameterizations of a nonlinear population model which includes density dependence
of population growth. For the linear model, let $E_i(t)$ 
again be an ecological variable, $i=1,2$.
We use AR(1) dynamics, 
$E_i(t+1)=\beta E_{i}(t)+ \sqrt{1-\beta^2} \epsilon_i(t)$, with $\beta=0.5$. 
The environmental noises $\epsilon_i(t)$ were 
standard-normal random variables that were independent for distinct times, $t$, but 
exhibited different kinds of 
dependence across locations in different simulations (see below).
The variable $E$ is general. It could represent deviations
of a population density from a carrying capacity, deviations of
total plant community biomass from an average value,
flux of a biogeochemical variable such as methane, or other quantities.
The nonlinear population model was a stochastic, multi-habitat-patch version 
of the Ricker model, 
$P_i(t+1)=P_i(t)\exp \left[r\left(1-P_i(t)/K  \right)+\sigma\epsilon_i(t)\right]$ for $i=1,2$, using $r=0.5$, 
$K=100$, $\epsilon_i(t)$
as above, and $\sigma=0.1$ or 
$\sigma=1$ in different simulations. 
When $\sigma=0.1$, population dynamics stay close to the carrying capacity, 
$K$, where the model equation can be well approximated by a linear equation,
and the nonlinearities of the model therefore have limited influence on dynamics. When
$\sigma=1$, model dynamics are strongly nonlinear because the stochastic 
component of the model causes populations to stray far from the carrying capacity. 
We refer to these as
the weak-noise and strong-noise cases, though the importance of the noise here
is that, when strong, it brings the nonlinearities of the model into play.

For each of the three model setups above (the linear model and the nonlinear model
with weak and strong stochasticity), for the Clayton and survival Clayton copula families, 
for each $\tau=0.1,0.2,\ldots,0.9$, and for each of
$50$ replicate simulations, we generated 5000 noise pairs 
$(\epsilon_1(t),\epsilon_2(t))$ from the
bivariate random variable with standard-normal marginals and with 
the given copula family and the given Kendall
correlation $\tau$.
We then used this noise to drive the model, and retained both the
noise and population values for the final 500 
time steps. For each simulation, the following
statistics were then computed for noises and populations: 
Pearson, Spearman and Kendall correlations, $\cor_l$, $\cor_u$, $\Ps_l$, $\Ps_u$, 
$\Dsq_l$, $\Dsq_u$, $\cor_l - \cor_u$, $\Ps_l - \Ps_u$, and $\Dsq_u - \Dsq_l$. 
Values were plotted against $\tau$ for noises and populations. If the 
hypothesis from the Introduction was reasonable that characteristics of 
the copula structure of spatial dependence in an
ecological variable may be inherited from characteristics of spatial 
dependence in an environmental variable through a Moran-like effect, then 
plots should be similar for populations and noises.

>>>>

Well-developed approaches do exist, however, and are 
applied widely in other fields, 
that provide a description of the association 
between two or more variables that is mathematically provably 
complete [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]. 
The approaches are based on 
the concept of the *copula*. Copula approaches separate information in a bivariate 
random variable into two parts: the information in the 
marginal distributions (which says nothing about the association between the 
variables), and the remaining information (which is solely about
the association). We here introduce some simple concepts, based on ranks,
that relate to copulas and give a conceptual 
flavor of copula ideas to be introduced 
formally in the "Background" section below (section \ref{Background}). Given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), information about the structure of the association
between $x$ and $y$ can be separated from information contained 
in the marginal distributions by 
considering the plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The rank
of the smallest element in a set is $1$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$; they relate to 
the empirical cumulative distribution function of the $x_i$ and $y_i$,
respectively. We henceforth abbreviate "cumulative distribution function" as "cdf". 
See, for instance,
Fig. \ref{fig_cop_pedag1}C-D, which show the normalized ranks of 
A-B. Ranking makes the marginal distributions of 
the component datasets uniform, isolating the 
dependence structure. Dependence structure and marginals can then be studied 
separately. Normalized rank plots such as Fig. \ref{fig_cop_pedag1}C-D
relate to copulas in a way to be described 
(section \ref{Background}). Note that Pearson correlation, even though it is the most commonly
used measure of association, is modified by monotonic
transformation of the component variables (Appendix 
\ref{SM-pedagog1_details}, Fig. \ref{SM-fig_transf_pearson_spearman}),
and therefore reflects not only dependence information, but also
information on the marginals [@Genest2007]. 
Spearman and Kendall correlations, being rank-based, 
are not influenced by monotonic transformations of the variables. 
They provide information solely on the dependence between those variables.

@@

A second reason why we may expect detailed understanding of dependence structures
to benefit ecological
research is that prior work demonstrated, using copulas, complex 
structure in the spatial dependence of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. An environmental variable 
measured through time in two locations may show strong tail associations between
the locations if
intense meteorological events are also widespread, as seems 
frequently to be the case: extreme values are associated with intense events,
so happen in both places at the same time, whereas moderate values of the 
environmental variable may be associated with local phenomena, which differ
between the locations.
Spatial dependence in an environmental variable 
tends to beget spatial dependence 
in fluctuations of populations or other ecosystem variables 
(this is called \emph{spatial synchrony}) influenced by the environmental 
variable. This is called the \emph{Moran effect}. 
If asymmetric tail associations, or other complex dependence structure, is transmitted 
from environmental to ecological variables, then we would expect complex dependence 
structure and tail associations to be a common feature of the 
spatial synchrony of population, community, 
biogeochemical and other environmentally influenced ecological variables.
Synchrony attracts major interest in 
ecology [@Liebhold2004; @Sheppard2016; @walteretal2017].

@@

We approached our overall goal of exploring the potential of copula approaches 
for ecology by addressing the following specific questions. (Q1) Do datasets
in ecology have dependence structure distinct from that of a multivariate 
Gaussian/normal
distribution (here called non-normal copula structure)? Do positively associated 
ecological variables show tail 
associations distinct from those of a normal distribution, and in particular
do they show asymmetric tail associations? Normal copula structure is assumed
by standard approaches that use multivariate 
normal distributions or distributions obtained by transforming
the marginals of a normal distribution. So Q1 asks whether ecological
data contain dependency information distinct from a standard or default case. 
(Q2) If the answer to
Q1 is "yes", then what are some possible causes/mechanisms of non-normal
copula structure and asymmetric tail associations in ecology? And (Q3), what are the 
consequences of non-normal copula
structure and asymmetric tail associations for ecological understanding and
applications? After section \ref{Background}, 'Background on copulas', we address Q1 
(sections \ref{Data} through \ref{Results} of the paper) 
by analyzing several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
We address Q2 (sections \ref{Causes} through \ref{Results_Q2}), using 
simple models. We address Q3 (sections \ref{Consequences} through \ref{Results_Q3}) using 
both data and models. Section \ref{Discussion} is the Discussion.
Multiple analyses are brought to bear on each question, and these
are summarized diagrammatically in Fig. \ref{MasterFigure}, which can 
serve as a *post hoc* guide to the paper. 
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000],
and have also been used effectively, but rarely so far, in ecology [@Valpine2014; 
@Anderson2018; @Popovic2018]. 
We argue that benefits of wider use of
copula descriptions of dependence in ecology will be substantial.

@@

Plots using normalized ranks were proposed in the Introduction as being conceptually
similar to copulas; we explain the connection. As stated above, the copula associated
with a random variable $(X,Y)$ is the cdf of the random variable
$(F_X(X),F_Y(Y))$. Given a sample $(x_i,y_i)$, $i=1,\ldots,n$, from $(X,Y)$, let $\hat{F}_X$ and
$\hat{F}_Y$ be the empirical cdfs associated with the $x_i$ and $y_i$, respectively.
These are step functions of $x$ and $y$, respectively, that start at $0$ for low values of 
$x$ and $y$ and jump by $1/n$
at each of the data points. Therefore $\hat{F}_X(x_j)$ is the rank of $x_j$ in
the set $\{x_i : i=1,\ldots,n\}$, divided by $n$, 
and analogously $\hat{F}_Y(y_j)$ is the rank of $y_j$ in the set 
$\{y_i : i=1,\dots,n\}$, divided
by $n$. For large $n$, $\hat{F}_X(x_i)$ and $\hat{F}_Y(y_i)$ 
approximate the cdfs $F_X$ and $F_Y$, respectively. 
But $\hat{F}_X(x_i)$ equals the normalized rank of $x_i$ times
$n/(n+1)$, and likewise $\hat{F}_Y(y_i)$ equals the normalized rank of $y_i$ times
$n/(n+1)$. So the normalized rank in turn approximates the empirical cdf and therefore 
the cdf. Thus the normalized rank pairs $(u_i,v_i)$ for $i=1,\ldots,n$ can be 
regarded as approximate samples from the random variable $(F_X(X),F_Y(Y))$, 
which is the random variable associated with the copula of $(X,Y)$.
The scatterplot of $v_i$ versus $u_i$ can be used to infer aspects of copula 
structure (see section \ref{Methods} below).

@@

Model selection methods give the *relative* support of several models, but do not 
indicate whether any of the models are an objectively good fit. To test that, 
we tested the goodness of fit of our AIC-best copula family using a 
bootstrapping procedure of @Wang2000 and @Genest2006,
implemented as `BiCopGofTest` in `VineCopula`. The procedure performed 
one test based on a Cramer-von Mises statistic and another based on a 
Kolmogorov-Smirnov statistic. To keep computation times reasonable, 
a run using $100$ bootstraps was performed. If
the $p$-value from either test was $<0.2$, tests were re-run with 
$1000$ bootstraps.

We fit 16 bivariate copula families, exhibiting a variety of 
lower- and upper-tail dependence characteristics, with bivariate datasets. The purpose
of using a large collection of families was to include a 
variety of alternative dependence structures to have a robust model
selection and multi-model inference procedure. For that purpose, it 
is not important for the reader to
understand the details of these copulas,
and, additionally, these copulas have been 
described in detail elsewhere [@Joe1997; @Brechmann2013]. So 
we only briefly identify each family, say a few words about its tail 
dependence, and provide pictorial descriptions (Figs 
\ref{figped_N}-\ref{figped_BB1}, \ref{SM-figped_SC} and \ref{SM-figped_G}-\ref{SM-figped_SBB8}). 
The pictorial descriptions can also be used to aid quick comparisons of copula
families when alternative families are being considered for future applications. 
We used several families that can exhibit positive lower-tail dependence 
(of strength depending on parameters) and 
zero upper-tail dependence: the Clayton, survival Gumbel, survival Joe
and survival BB6 families. We used several families that can exhibit positive
upper-tail dependence (strength depends on parameters) and zero lower-tail
dependence: the survival Clayton, Gumbel, Joe and BB6 families. We used
families that show zero upper- and lower-tail dependence: the normal
and Frank families. These families have pdfs symmetric about the line
$v=-u+1$. We also used several families that can show both upper- and
lower-tail dependence, in relative amounts depending on parameter
values: the BB1, survival BB1, BB7 and survival BB7 families. 
We also used the BB8 family, which shows zero lower-tail dependence,
and zero upper-tail dependence except for a boundary case for the parameters.
And we used the survival BB8 copula, which shows zero upper-tail dependence, 
and zero lower-tail dependence except for a boundary case.
"Survival" families are rotations of the copula with the similar name by
180 degrees. 
See @Joe1997 and @Brechmann2013 for details on all families. We used the implementations
of provided in the `VineCopula` package for the R programming
language. See Figs \ref{figped_N}-\ref{figped_BB1}, \ref{SM-figped_SC} and \ref{SM-figped_G}-\ref{SM-figped_SBB8} 
for visual depictions of the pdfs of these copulas and how
pdfs and tail dependence are influenced by the parameters. See 
Table \ref{tab_copula_info} for a summary.

@@

For large datasets (large $n$),
we used $l_b$ and $u_b$ close together without incurring undue
sampling variation in our statistics, and we considered 
multiple bands $(l_b,u_b)$ to understand how association
varies across different parts of the distributions. But for datasets with 
smaller $n$ we considered only $l_b=0$, $u_b=0.5$ and $l_b=0.5$, $u_b=1$,
abbreviating $\cor_l=\cor_{0,0.5}$ ($l$ is for "lower") 
and $\cor_u=\cor_{0.5,1}$ ($u$ is for "upper"). Likewise $\Ps_l=\Ps_{0,0.5}$,
$\Ps_u=\Ps_{0.5,1}$, $\Dsq_l=\Dsq_{0,0.5}$, $\Dsq_u=\Dsq_{0.5,1}$.
To test for asymmetry of association in upper and lower portions of distributions,
we used differences $\cor_l-\cor_u$, $\Ps_l-\Ps_u$ and $\Dsq_u-\Dsq_l$
for smaller datasets (note the opposite order
in the last of these); and $\cor_{0,u_b}-\cor_{1-u_b,1}$, 
$\Ps_{0,u_b}-\Ps_{1-u_b,1}$ and
$\Dsq_{1-u_b,1}-\Dsq_{0,u_b}$ with $u_b$ close 
to $0$ for large datasets.
We tested our statistics in Appendix \ref{SM-Test_npa_stat} (see also 
Figs \ref{SM-fig_stat_testing35}-\ref{SM-fig_stat_testing1000}).

@@

The first mechanism relates to the ideas in the Introduction about 
Liebig's law of the minimum, and to nonlinear influences of environmental
variables on ecological variables.
If an environmental variable influences an ecological variable 
disproportionately in one of its tails, we explored whether the ecological
variable could then exhibit asymmetric tail associations across space.
Let $E_i(t)$ be an ecological variable 
measured at location $i$ ($i=1,2$) at time $t$.
Assume the dynamics 
$E_i(t+1)=bE_i(t)+g(\varepsilon_i(t))+a \delta_i(t)$, where 
the $\delta_i(t)$ are standard-normally distributed 
and independent across time and locations, $a=0.2$,
$b=0.1, -0.1, 0.5, -0.5$ in different simulations, and 
the $(\varepsilon_1(t),\varepsilon_2(t))$ were drawn, independently through
time, from a bivariate normal distribution with $\var(\varepsilon_i)=1$ and 
$\cov(\varepsilon_1,\varepsilon_2)=0.8$. Thus basic ecological dynamics follow the
very simple AR(1) formulation, influenced by a "regional" environmental factor which
is correlated across locations ($\varepsilon$) and by a local factor ($\delta$). 
The function $g$ describes how $\varepsilon_i(t)$ influences
$E_i(t)$. We used $g$ equal to $g_1$ or $g_2$ in different simulations:
$g_1(\varepsilon)$ equals $\varepsilon$ if $\varepsilon<0$ and $0$ otherwise;
$g_2(\varepsilon)$ equals $\varepsilon$ if $\varepsilon>0$ and $0$ otherwise. Thus
$g_1$ represents environmental effects that negatively impact populations,
but only below the threshold $\varepsilon=0$; 
and $g_2$ represents effects that positively impact populations, but 
only above $\varepsilon=0$.
The values of $b$ provide a modicum of exploration of whether ecological 
dynamics may also influence how complex dependencies between
the $E_i(t)$ arise: 
negative $b$ corresponds to overcompensating dynamics 
and positive $b$ to undercompensating dynamics; larger $|b|$
means slower return to equilibrium after a disturbance.

For each $b$ and $g_i$, we simulated the model for $25000$ time
steps and retained the $E_i(t)$ for the final $2500$ time 
steps. We applied our 
nonparametric statistic $\cor_{0,0.1}-\cor_{0.9,1}$ 
and our Spearman- or Kendall-preserving normal-copula 
surrogate comparison methods (see section \ref{Methods}) to these 
outputs to discover if the model could
produce asymmetric tail associations (and therefore non-normal copula structure)
between the $E_i(t)$. Because $(\varepsilon_1,\varepsilon_2)$
and $(\delta_1,\delta_2)$ have normal copula structure (they were drawn
from bivariate normal distributions), the Moran 
mechanism analyzed below does not operate here.

Our second mechanism is an extension of the well known Moran
effect, and was summarized conceptually in the Introduction. 
We consider a linear model, as well as 
two parameterizations of a nonlinear population model which includes density dependence
of population growth. For the linear model, let $E_i(t)$ 
again be an ecological variable, $i=1,2$.
We use AR(1) dynamics, 
$E_i(t+1)=\beta E_{i}(t)+ \sqrt{1-\beta^2} \varepsilon_i(t)$, with $\beta=0.5$. 
The environmental noises $\varepsilon_i(t)$ were 
standard-normal random variables that were independent for distinct times, $t$, but 
exhibited different kinds of 
dependence across locations in different simulations (see below).
The variable $E$ is general. It could represent deviations
of a population density from a carrying capacity, deviations of
total plant community biomass from an average value,
flux of a biogeochemical variable such as methane, or other quantities.
The nonlinear population model was a stochastic, multi-habitat-patch version 
of the Ricker model, 
$P_i(t+1)=P_i(t)\exp \left[r\left(1-P_i(t)/K  \right)+\sigma\varepsilon_i(t)\right]$ for $i=1,2$, using $r=0.5$, 
$K=100$, $\varepsilon_i(t)$
as above, and $\sigma=0.1$ or 
$\sigma=1$ in different simulations. 
When $\sigma=0.1$, population dynamics stay close to the carrying capacity, 
$K$, where the model equation can be well approximated by a linear equation,
and the nonlinearities of the model therefore have limited influence on dynamics. When
$\sigma=1$, model dynamics are strongly nonlinear because the stochastic 
component of the model causes populations to stray far from the carrying capacity. 
We refer to these as
the weak-noise and strong-noise cases, though the importance of the noise here
is that, when strong, it brings the nonlinearities of the model into play.

For each of the three model setups above (the linear model and the nonlinear model
with weak and strong stochasticity), for the Clayton and survival Clayton copula families, 
for each $\tau=0.1,0.2,\ldots,0.9$, and for each of
$50$ replicate simulations, we generated 5000 noise pairs 
$(\varepsilon_1(t),\varepsilon_2(t))$ from the
bivariate random variable with standard-normal marginals and with 
the given copula family and the given Kendall
correlation $\tau$.
We then used this noise to drive the model, and retained both the
noise and population values for the final 500 
time steps. For each simulation, the following
statistics were then computed for noises and populations: 
Pearson, Spearman and Kendall correlations, $\cor_l$, $\cor_u$, $\Ps_l$, $\Ps_u$, 
$\Dsq_l$, $\Dsq_u$, $\cor_l - \cor_u$, $\Ps_l - \Ps_u$, and $\Dsq_u - \Dsq_l$. 
Values were plotted against $\tau$ for noises and populations. If the 
hypothesis from the Introduction was reasonable that characteristics of 
the copula structure of spatial dependence in an
ecological variable may be inherited from characteristics of spatial 
dependence in an environmental variable through a Moran-like effect, then 
plots should be similar for populations and noises.

The next mechanism we investigated is evolutionary, and pertains to bivariate 
trait data across species, e.g., our bird and mammal 
data. This mechanism is a hypothetical explanation for the bias toward right-tail association
observed in those data (Fig. \ref{fig_biv_multi_raw_cop}F,G). 
The hypothesis is that asymmetric tail association occurs in evolutionary changes
in bivariate characters, and gives rise to asymmetric tail association between the 
two character values across extant species.
We simulated bivariate character evolution on an estimate of 
the phylogeny, taken from @GenoudIM2018, of $817$ mammal species.
The root character state and change 
across each branch were randomly chosen from matrices of one million 
independent draws from bivariate distributions showing one of five distinct types of copula 
structure: 1) extreme or 2) moderate left-tail dependence,
3) symmetric tail dependence, or 4) moderate or 5) extreme
right-tail dependence (Appendix \ref{SM-evol_mech}). 
All distributions had standard-normal marginals and Spearman 
correlation $0.875$ between components, so our simulations assess 
the impact of copula structure only. For 
each of the five copulas, mammalian character
evolution was simulated 100 times. For each simulation, symmetry of tail associations 
of the two characters across phylogeny tips was assessed using our nonparametric
statistics.
We hypothesized that cases 1 and 2 above would yield stronger left- than right-tail 
associations in tip characters, and cases 4 and 5 would yield the reverse. 
The simulator was written in Python and used version 4.4 of the DendroPy 
package [@DendroPy].
